Compositional Quantitative Resource Analysis of C Programs

### 2014-06-25 paper/main.tex
description for the rules

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of the existing techniques in practice.
\begin{itemize}
\item Analysis tools are black boxes that either deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\item Existing techniques often rely on complex external tools such as
  abstract interpretation based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
%
While there has been lot of progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-3ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
%
%
In this paper, we present a novel resource analysis framework for C programs
that is based on a solid semantic foundation. It works automatically,
efficiently and precisely for many examples, can be used interactively
for manual bound derivation if the automation fails, and produces
verifiable certificates for the bounds.  Our contributions are as
follows.
%
\begin{enumerate}
\item We define a operational cost semantics for C programs that
  accounts a constant cost for each statement.  The cost is parametric
  in a cost metric and the semantics models cost of terminating and
  non-terminating executions.  A specific \emph{tick} function can be
  inserted in the source code to further customize the cost model.

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q \implies \code{istrue}~\sem{e}_\state + M_a }
  { \code{assert}~e }
  { Q }
}
%
\and
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q }{ \code{skip} }{ Q }
}
%
\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}
%
\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { R\,(\state(x)) }{ \code{return}~x }{ Q }
}
%
\and
\Rule{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}
%
\and
\Rule{L:Call}
{
  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
  P \models P_f \, y \, (\state(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q - M_r }
}
%
\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}
%
\and
\Rule{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}
%
\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { P \land \code{istrue}~\sem{e}_\state - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { P \land \code{isfalse}~\sem{e}_\state - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}
%
\and
\Rule{L:Weaken}
{
  P \models P' \\
  \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \\
  Q' \models Q \\
  B' \models B \\
  \forall v.\, (R'\,v \models R\,v)
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \mathbb Q_0^+
}
{ \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x }
}
%
\and
\Rule{L:Extend}
{ \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
\\
  \forall f\,P_f\,Q_f.\,
  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\end{mathpar}

@@

\paragraph{Semantic meaning of triples.}

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{Q:Skip}
{}
{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
%
\and \Rule{Q:Break}
{}
{  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') }
%
\and \Rule{Q:Return}
{ P = Q_R[\Vret/x]  
\\ \forall i \in \dom{P} . \, p_i = q_i
}
{ B; (\Gamma,Q_R); (\Gamma, Q{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
%
\and \Rule{Q:Assert}
{}
{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
%
\and \Rule{Q:Loop}
{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
%
\and \Rule{Q:Set}
{ q'_{xy}, q'_{yx} \in \mathbb Q_0^+
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ B; R; (\Gamma[x/y], Q {+} M_s) \vdash x \gets y \dashv (\Gamma, Q') }
%
\and \Rule{Q:IncrPos}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
\\\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx}
}
{ \textstyle
  B; R; (\Gamma[x/x {+} y], Q {+} M_s) \vdash x \gets x + y \dashv (\Gamma, Q')
}
%
\and \Rule[rightskip=1cm]{Q:IncrNeg}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
\\\\ \textstyle q'_{y0} = q_{y0}
      + \sum_{u \in \mathcal U} q_{ux}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {-} y],Q {+} M_s) \vdash x \gets x - y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:Incr}
{ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x \pm y], Q {+} M_s) \vdash x \gets x \pm y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:If}
{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
}
{ B; R; (\Gamma, Q) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
%
\and\Rule{Q:Seq}
{  B; R; (\Gamma, Q {-} M_s^1) \vdash S_1 \dashv (\Gamma', Q')
\\ B; R; (\Gamma', Q' {-} M_s^2) \vdash S_2 \dashv (\Gamma'',Q'')
}
{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
%
\and\Rule{Q:Call}
{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f)
\and \text{Loc} = \text{Locals}(Q)
\and \forall i \neq j . \, x_i \neq x_j
\and c \in \mathbb Q_0^+
\\ Q = P + S
\and Q' = P' + S
\and U = Q_f[\Vargs / \vec x]
\and U' = Q'_f[\Vret/r]
\\ \forall i \in \dom{U} . \, p_i = u_i
\and \forall i \in \dom{U'} .\, p'_i = u'_i 
\and \forall i \not\in \dom{U'} .\, p'_i = 0
\and \forall i \not\in \text{Loc}
  .\, s_i = 0
}
{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c)
  \vdash r \gets f(\vec x) \dashv
  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c)
}
%
\and \Rule{Q:Weak}
{ \Gamma_1 \models \Gamma_2
\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
\\ \Gamma'_2 \models \Gamma'_1
}
{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
%
\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |\inter x y| \}
\and \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |\inter x y| \le u_{xy} \}
\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
}
{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
%
\and \Rule{Q:Abst}
{ \fenv f = (\vec y, S_f)
\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S \dashv (\Gamma', Q')
}
{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
%
\end{mathpar}
\caption{Inference rules of the quantitative analysis}
\label{fig:proof}
\end{figure*}

>>>>

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of the existing techniques in practice.
\begin{itemize}
\item Analysis tools are black boxes that either deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\item Existing techniques often rely on complex external tools such as
  abstract interpretation based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
%
While there has been much progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-3ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
%
%
In this paper, we present a novel resource analysis framework for C programs
that is based on a solid semantic foundation. It works automatically,
efficiently and precisely for many examples, can be used interactively
for manual bound derivation if the automation fails, and produces
verifiable certificates for the bounds.  Our contributions are as
follows.
%
\begin{enumerate}
\item We define a operational cost semantics for C programs that
  accounts a constant cost for each statement.  The cost is parametric
  in a cost metric and the semantics models cost of terminating and
  non-terminating executions.  A specific \emph{tick} function can be
  inserted in the source code to further customize the cost model.

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q \implies \code{istrue}~\sem{e}_\state + M_a }
  { \code{assert}~e }
  { Q }
}
%
\and
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q }{ \code{skip} }{ Q }
}
%
\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}
%
\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { R\,(\state(x)) }{ \code{return}~x }{ Q }
}
%
\and
\Rule{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}
%
\and
\Rule{L:Call}
{
  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
  P \models P_f \, y \, (\state(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q - M_r }
}
%
\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}
%
\and
\Rule{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}
%
\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { P \land \code{istrue}~\sem{e}_\state - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { P \land \code{isfalse}~\sem{e}_\state - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}
%
\and
\Rule{L:Tick}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q {+} M_t(n) }
  { \code{tick}(n) }
  { Q }
}
%
\and
%
\Rule{L:Weaken}
{
  P \models P' \\
  \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \\
  Q' \models Q \\
  B' \models B \\
  \forall v.\, (R'\,v \models R\,v)
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \mathbb Q_0^+
}
{ \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x }
}
%
\and
\Rule{L:Extend}
{ \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
\\
  \forall f\,P_f\,Q_f.\,
  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\end{mathpar}

@@

\paragraph{Rules of the Quantitative Logic}

\pref{fig:logic} shows the inference rules of the quantitative logic.
The rules are slightly simplified in comparison to the implemented
rules in Coq.  The main difference is that the presented version do
not formalize an auxiliary state that is needed to relate pre and
postconditions (see the following paragraph for more details).

In the rule {\sc L:Skip}, we do not have to account for any resource
consumption.  As a result, the precondition $Q$ can be any (potential)
function and we only have to make sure that we do not end up with more
potential.  Since the execution of \code{skip} leaves the program
state unchanged, we can simply use the precondition as postcondition.
The potential functions $B$ for the \code{break} and $R$ for the
\code{return} part of the postcondition are not reachable and can
therefore be arbitrary.

The rule {\sc L:Assert} is strange XXX.

The rules {\sc L:Break} and {\sc L:Return} are similar to the rule
{\sc L:Skip}.  The only difference is that we replace requirement of
the \code{skip} part of the postcondition with analogue requirements
on the \code{break} and \code{return} part, respectively.  The
aforementioned rules not trigger any stack-space consumption nor
release of stack-space.  This might seem strange in the \code{return}
case but will become clear in the description of the {\sc L:Call} rule
later on.

The rules {\sc L:AssignL} and {\sc L:AssignC} for local and global
assignment follow the traditional backward style of Hoare logic. It is
ensured that the precondition takes the same value on the initial
state as the \code{skip} part of the postcondition on the state
modified by assignment.

Despite its seemingly simplicity, the {\sc L:Seq} rule must not be
overlooked to understand how the quantitative Hoare logic works.  We
have to define it in such a way that it accounts for early
exits in statements. For instance, if $S_1$ contains a \code{break}
statement then $S_2$ will never be executed so we must ensure in the
\code{break} part of $S_1$'s postcondition that the \code{break} part
of $S_1;S_2$ holds. For the same reason, the \code{return} part of
$S_1$'s postcondition is special.

The {\sc L:Loop} rule uses the same principles as the {\sc Seq} rule
to tweak the final postcondition. In the case of {\sc L:Loop}, we
simply ensure that the \code{break} part of the inner statement
becomes the \code{skip} part of the overall statement. We use $\bot$
as the \code{break} part of the $\code{loop}\, S$ statement since its
operational semantics prevent it from terminating differently than by
a \code{skip} or a \code{return}.

The {\sc L:Call} rule accounts for the actual stack-space usage of
programs. It enforces that enough stack space is available to call the
function $f$ by adding $M(f)$ to the precondition and the
postcondition. The pre- and postconditions are taken from the context
$\Gamma$. This context is extended using the {\sc L:Abstract} rule
described below. The assertions in the context are parametric with
respect to both the function argument value and the return value. This
allows to specify a bound for a function whose recursion depth depends
on an input parameter.  The argument parameter is instantiated by the
call rule using the result of the evaluation of the argument
expression in the current state. Note the symmetry of the rule {\sc
  L:Call}: $M(f)$ is added on both sides to account for the stack
space that becomes available after the call.  This justifies that the
{\sc L:Return} rule does not account for stack-space release.

Finally, we describe the rules which are not syntax directed. The rule
{\sc L:Abstract} allows to make a proof on any statement with an extra
hypothesis in the derivation context provided that we have a proof
that this hypothesis is true. We can see as the quantitative
interpretation of the usual logical Modus Ponens.

There are two weakening rules available in the quantitative Hoare
logic.  The framing rule {\sc L:Frame} is designed to weaken a
statement by stating that if $S$ needs $P$ bytes to run and leaves $Q$
bytes free at its end, then it can very well run with $P + c$ bytes
and return $Q + c$ bytes. It comes very handy when we want to prove
tight bounds using the $\max$ function as demonstrated
in~\pref{fig:example_derivation}. The consequence {\sc L:Conseq} rule
is directly imported from classical Hoare logics except that instead
of using the logical implication $\Rightarrow$ we use the quantitative
$\ge$. This rule indeed weakens the statement since it requires more
resource to run the statement and yields less than what has been
proved to be available after its termination.

\paragraph{Auxiliary State and Stack Framing}

% As mentioned, our account of the logic in this article is slightly
% simplified compared to the Coq development to improve readability.
% % However, the described logic is very close to the implemented version.

% The main difference between the implemented logic and the logic
% described here is that the latter does not have an
% \emph{auxiliary state}.
% %
% Auxiliary state is a classic extension of Hoare logic (see for
% example~\cite{Nipkow02}).
%  It is used to share information
% between the pre- and postcondition of a triple.  In a logic without
% auxiliary state (or similar techniques) it is not possible to relate
% program states before and after a statement.  For example, you cannot
% specify that the function {\tt\ int\ twice\ ()\ $\{$i = i+i;$\}$\ }
% doubles the value of the variable $i$.  With an auxiliary variable $Z$
% it is possible specify this fact in Hoare logic using the triple $\htriple{i =
%   Z}{\code{twice()}}{ i = 2 \cdot Z}$.

% In the Coq implementation,
% assertions have the type $\State \times \mathit{Aux} \to \N \cup
% \{\infty\}$.  The auxiliary state is an arbitrary type in Coq and can
% be instantiated by the user.  Most of the rules of the logic remain
% unchanged in the presence of auxiliary state.  The
% only exception is the consequence rule \textsc{L:Conseq} that reads as
% follows.
% {  \footnotesize
% \begin{mathpar}
%   \RuleNolabel
%  {  \htriple{P'}{S}{R}
%  \\  
%  {\!\!\!\! \begin{array}{r}
% \forall \sigma \, a. P(\sigma,a) {\in} \N \! \implies  \! \exists a'.  
%      P(\sigma,a) {\geq} P'(\sigma,a') \hspace{3.3em} \\ \land (\forall \sigma'\, i.  R^i(\sigma',a') {\geq} Q^i(\sigma',a))
%  \end{array}}\!\!\!\!
%  }
%  { \htriple{P}{S}{Q}
%  }
% \end{mathpar}}
% This is the quantitative version of a consequence rule that has been
% introduced in the context of Hoare logic and is attributed to Martin
% Hofmann~\cite{Nipkow02}.  Its typical use case is when we apply the rule {\sc L:Call}
% to a recursive call.  In this case, the Hoare triple for the function
% call is proved by an assumption from the derivation context with
% a slightly different auxiliary state.  In the example derivation
% in~\pref{fig:bsearch_deriv} this different state is $Z - 1$.  Adapting
% the derivation hypothesis to prove the recursive call is enabled in
% our logic by the extended consequence rule introduced above.

% \paragraph{Stack Framing}

% A minor difference in the function application rule is that we only
% present the rule for function calls with a single argument and without
% \emph{framing of stack assertions}.  The latter is necessary in the
% code of a caller to carry over information on the local environment
% from the precondition of the function call to the postcondition of the
% function call.  This is a general problem in Hoare logic and stack
% framing is a well known solution.  For instance, we would like to prove
% something like $\htriple{y}{ f(x) }{y}$ where $y$ is a local variable.
% This is not possible with the rule \textsc{L:Call} but it is with the more
% general rule in the Coq implementation~\iffullversion{\cite{coqfiles}}{\cite{anoncoq}}.

\paragraph{Soundness of Quantitative Hoare Triples.}

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{Q:Skip}
{}
{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
%
\and \Rule{Q:Break}
{}
{  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') }
%
\and \Rule{Q:Return}
{ P = Q_R[\Vret/x]  
\\ \forall i \in \dom{P} . \, p_i = q_i
}
{ B; (\Gamma,Q_R); (\Gamma, Q{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
%
\and \Rule{Q:Assert}
{}
{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
%
\and \Rule{Q:Loop}
{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
%
\and \Rule{Q:Set}
{ q'_{xy}, q'_{yx} \in \mathbb Q_0^+
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ B; R; (\Gamma[x/y], Q {+} M_s) \vdash x \gets y \dashv (\Gamma, Q') }
%
\and \Rule{Q:IncrPos}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
\\\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx}
}
{ \textstyle
  B; R; (\Gamma[x/x {+} y], Q {+} M_s) \vdash x \gets x + y \dashv (\Gamma, Q')
}
%
\and \Rule[rightskip=1cm]{Q:IncrNeg}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
\\\\ \textstyle q'_{y0} = q_{y0}
      + \sum_{u \in \mathcal U} q_{ux}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {-} y],Q {+} M_s) \vdash x \gets x - y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:Incr}
{ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x \pm y], Q {+} M_s) \vdash x \gets x \pm y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:If}
{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
}
{ B; R; (\Gamma, Q) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
%
\and\Rule{Q:Seq}
{  B; R; (\Gamma, Q {-} M_s^1) \vdash S_1 \dashv (\Gamma', Q')
\\ B; R; (\Gamma', Q' {-} M_s^2) \vdash S_2 \dashv (\Gamma'',Q'')
}
{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
%
\and\Rule{Q:Call}
{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f)
\and \text{Loc} = \text{Locals}(Q)
\and \forall i \neq j . \, x_i \neq x_j
\and c \in \mathbb Q_0^+
\\ Q = P + S
\and Q' = P' + S
\and U = Q_f[\Vargs / \vec x]
\and U' = Q'_f[\Vret/r]
\\ \forall i \in \dom{U} . \, p_i = u_i
\and \forall i \in \dom{U'} .\, p'_i = u'_i 
\and \forall i \not\in \dom{U'} .\, p'_i = 0
\and \forall i \not\in \text{Loc}
  .\, s_i = 0
}
{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c)
  \vdash r \gets f(\vec x) \dashv
  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c)
}
%
\Rule{Q:Tick}
{}
{ B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
%
\and \Rule{Q:Weak}
{ \Gamma_1 \models \Gamma_2
\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
\\ \Gamma'_2 \models \Gamma'_1
}
{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
%
\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |\inter x y| \}
\and \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |\inter x y| \le u_{xy} \}
\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
}
{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
%
\and \Rule{Q:Abst}
{ \fenv f = (\vec y, S_f)
\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S \dashv (\Gamma', Q')
}
{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
%
\end{mathpar}
\caption{Inference rules of the quantitative analysis}
\label{fig:proof}
\end{figure*}
